---
title: "NASA Meteorite Landings Classification"
author: "Eno Umoren, Camila Nunez Polanco, and Rohil Nataraj"
date: "2026-04-23"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(caret)
library(pROC)

set.seed(123)
theme_set(theme_minimal())
```

## Research question

Which factors--composition, mass, year, and geographic location--are associated with the probability that a meteorite is classified as `Fell` rather than `Found`?

## Load and clean the data

```{r load-data}
data_url <- "https://data.nasa.gov/docs/legacy/meteorite_landings/Meteorite_Landings.csv"
meteorite_raw <- read.csv(data_url, stringsAsFactors = FALSE)

meteorite <- meteorite_raw |>
  filter(
    nametype == "Valid",
    !is.na(mass..g.),
    !is.na(fall),
    !is.na(year),
    !is.na(recclass),
    !is.na(reclat),
    !is.na(reclong),
    year >= 860,
    year <= 2016,
    reclong >= -180,
    reclong <= 180,
    reclat >= -90,
    reclat <= 90,
    reclat != 0 | reclong != 0
  ) |>
  rename(mass = mass..g.) |>
  mutate(
    composition = case_when(
      grepl("Iron", recclass, ignore.case = TRUE) ~ "Iron",
      grepl("Mesosiderite|Pallasite", recclass, ignore.case = TRUE) ~ "Stony-Iron",
      TRUE ~ "Stone"
    ),
    mass_group = case_when(
      mass < 100 ~ "Small (<100 g)",
      mass < 10000 ~ "Medium (100-10,000 g)",
      TRUE ~ "Large (>10,000 g)"
    ),
    year_group = if_else(year <= 1986, "Early (860-1986)", "Later (1987-2013)"),
    region = case_when(
      reclat > 0 & reclong < 0 ~ "Northwest",
      reclat > 0 & reclong >= 0 ~ "Northeast",
      reclat <= 0 & reclong < 0 ~ "Southwest",
      TRUE ~ "Southeast"
    ),
    fall = factor(fall, levels = c("Found", "Fell")),
    composition = factor(composition),
    mass_group = factor(
      mass_group,
      levels = c("Large (>10,000 g)", "Medium (100-10,000 g)", "Small (<100 g)")
    ),
    year_group = factor(year_group, levels = c("Early (860-1986)", "Later (1987-2013)")),
    region = factor(region, levels = c("Northeast", "Northwest", "Southeast", "Southwest"))
  ) |>
  select(name, fall, mass, mass_group, composition, year, year_group, region)

cat("Raw observations:", nrow(meteorite_raw), "\n")
cat("Clean observations:", nrow(meteorite), "\n")
round(prop.table(table(meteorite$fall)), 4)
```

## Exploratory analysis

```{r composition-plot, fig.width=7, fig.height=4}
ggplot(meteorite, aes(composition, fill = fall)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Found" = "#1D4E7A", "Fell" = "#E63946")) +
  labs(
    title = "Fell versus Found by Meteorite Composition",
    x = "Composition",
    y = "Proportion",
    fill = "Status"
  )
```

```{r mass-plot, fig.width=7, fig.height=4}
ggplot(meteorite, aes(mass_group, fill = fall)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("Found" = "#1D4E7A", "Fell" = "#E63946")) +
  labs(
    title = "Fell versus Found by Meteorite Mass",
    x = "Mass category",
    y = "Proportion",
    fill = "Status"
  )
```

## Chi-square tests

```{r chi-square-tests}
predictors <- c("mass_group", "composition", "year_group", "region")

chi_square_results <- lapply(predictors, function(variable) {
  test <- chisq.test(table(meteorite[[variable]], meteorite$fall))
  data.frame(
    predictor = variable,
    chi_square = unname(test$statistic),
    degrees_of_freedom = unname(test$parameter),
    p_value = test$p.value,
    minimum_expected_count = min(test$expected)
  )
}) |>
  bind_rows()

chi_square_results
```

## Training and test sets

```{r split-data}
train_index <- createDataPartition(meteorite$fall, p = 0.70, list = FALSE)
train_data <- meteorite[train_index, ]
test_data <- meteorite[-train_index, ]

round(prop.table(table(train_data$fall)), 4)
round(prop.table(table(test_data$fall)), 4)
```

## Logistic regression models

```{r fit-models}
full_model <- glm(
  fall ~ mass_group + composition + region + year_group,
  data = train_data,
  family = binomial()
)

reduced_model <- glm(
  fall ~ mass_group + region,
  data = train_data,
  family = binomial()
)

interaction_model <- glm(
  fall ~ mass_group * region + composition + year_group,
  data = train_data,
  family = binomial()
)

model_comparison <- data.frame(
  model = c("Full", "Reduced", "Interaction"),
  residual_deviance = c(
    deviance(full_model),
    deviance(reduced_model),
    deviance(interaction_model)
  ),
  AIC = c(AIC(full_model), AIC(reduced_model), AIC(interaction_model))
)

model_comparison
```

## Test-set performance

```{r evaluate-models}
test_probabilities <- list(
  Full = predict(full_model, newdata = test_data, type = "response"),
  Reduced = predict(reduced_model, newdata = test_data, type = "response"),
  Interaction = predict(interaction_model, newdata = test_data, type = "response")
)

roc_objects <- lapply(
  test_probabilities,
  function(probability) roc(test_data$fall, probability, levels = c("Found", "Fell"), quiet = TRUE)
)

auc_results <- data.frame(
  model = names(roc_objects),
  AUC = vapply(roc_objects, function(object) as.numeric(auc(object)), numeric(1))
)

auc_results
```

```{r roc-plot, fig.width=7, fig.height=5}
plot(roc_objects$Full, col = "#2455A4", lwd = 3, main = "ROC Curve Comparison")
plot(roc_objects$Reduced, col = "#D1495B", lwd = 3, add = TRUE)
plot(roc_objects$Interaction, col = "#F4A261", lwd = 3, add = TRUE)
abline(a = 0, b = 1, lty = 2, col = "gray60")
legend(
  "bottomright",
  legend = paste0(auc_results$model, ": AUC = ", round(auc_results$AUC, 3)),
  col = c("#2455A4", "#D1495B", "#F4A261"),
  lwd = 3,
  bty = "n"
)
```

## Odds ratios for the full model

```{r odds-ratios}
odds_ratios <- data.frame(
  term = names(coef(full_model)),
  odds_ratio = exp(coef(full_model)),
  lower_95 = exp(confint.default(full_model)[, 1]),
  upper_95 = exp(confint.default(full_model)[, 2])
)

odds_ratios
```

## Confusion matrices

Because the outcome is highly imbalanced, a fixed 0.50 threshold can obscure performance on the minority `Fell` class. The code below uses each model's test-set ROC threshold that maximizes sensitivity plus specificity.

```{r confusion-matrices}
confusion_results <- lapply(names(roc_objects), function(model_name) {
  threshold <- as.numeric(coords(
    roc_objects[[model_name]],
    x = "best",
    best.method = "youden",
    ret = "threshold",
    transpose = FALSE
  ))

  predicted <- factor(
    ifelse(test_probabilities[[model_name]] >= threshold, "Fell", "Found"),
    levels = c("Found", "Fell")
  )

  list(
    model = model_name,
    threshold = threshold,
    confusion_matrix = confusionMatrix(predicted, test_data$fall, positive = "Fell")
  )
})

for (result in confusion_results) {
  cat("\n", result$model, "model; threshold =", round(result$threshold, 4), "\n")
  print(result$confusion_matrix)
}
```

## Limitations

- The classes are strongly imbalanced, with far more `Found` than `Fell` observations.
- Recorded locations reflect discovery and reporting patterns as well as meteorite processes.
- Geographic quadrants are broad and may hide important regional differences.
- The analysis is observational and does not establish causal relationships.

